Automated curriculum learning for Multi-Domain NMT by using loss-driven reward

### 2021-05-05 2021/emnlp/auto_cl_mdmt.tex
scores, second script

\title{Improving Multi-Domain Neural Machine Translation with Differential Data Selection}

@@

\begin{document}
\maketitle
\setlength{\abovedisplayskip}{2pt}
\setlength{\belowdisplayskip}{2pt}
\begin{abstract}
  When building machine translation systems, one often needs to make the best out of heterogeneous sets of parallel data in training to achieve the best test performance in one or several domain(s) of interest. This multi-source / multi-domain adaptation problem is often approached with instance selection or reweighting strategies, most of which pre-suppose an ex-ante assessment of the relevance of training instances. In this paper, we study the recently proposed Differential Data Selection (DDS) model and explore its ability to serve as a generic framework for these various situations. Our experiments study both domain adaptation and multi-domain learning and show that DDS often enables to outperform more heuristic adaptation strategies. We also introduce variants that boosts DDS performance of adapter-based multi-domain systems.\fyTodo{Keep this part ?}
%   This is a well-known issue, which has given raise to a rich litterature in under the umbrellas of domain adaptation or multi-domain

%   The priority of the in-domain performance in the overall evaluation affects the choice of choosing data. The problem falls in the data selection category in the domain adaptation topic. Several data selection methods pre-compute the domain-relatedness of training examples and use these scores to build mini-batches in the training while other works propose dynamic strategies to build training mini-batches based on reinforcement learning.
% In this study, we first formulate the most general algorithm for training the Neural Machine Translation (NMT) model given a predefined priority of the domains. Then, we apply a recently proposed method to solve the problem. We also report a disadvantage of DDS and fix it by proposing a minor development for DDS. Our experiments with a large sample of multi-domain systems show several important benefits of DDS in multi-domain NMT.
\end{abstract}

@@

\fyTodo{Label or covariate shift?} \revision{Answer: Covariate shift happens when D(x) differs between train and test; label shift happens when the conditional distribution $D(y|x)$ differs between train and test} 
This problem is often approached with \emph{static} instance selection or re-weighting strategies, where the available training data is used in proportion to its relevance for the chosen test conditions. Finding the optimal balance of training data is however a challenging task due, for instance, to similarity between domains / languages, but also due to regularization effects of out-of-domain data; it may also be suboptimal, as some target domains or languages might be easier to train than others. Several recent proposals \cite{Wang17instance,Zhang19curriculum,Kumar19reinforcement,Wang20learning-multi} have explored ways to instead consider more \emph{dynamic} data selection and sampling strategies. We study in this contribution the proposal of \citet{Wang20balancing}, initially introduced for multilingual MT, with the goal to evaluate its potential in various single-domain and multi-domain adaptation settings and assess DDS, both in its basic version and with our improvements, as an effective replacement to standard heuristics approaches.

Based on experimental results obtained on a diverse set of domains, our main conclusions are that (a) using DDS often yields overall performance that are as good as the standard fine-tuning strategy for domain adaptation; (b) DDS can effectively handle a variety of test target distributions without any meta-parameter search; (c) our extension of DDS is able to boost its performance when used in conjunction to the adapter model of \citet{Bapna19simple}.

% one often needs to make the best out of heterogeneous sets of parallel data in training to achieve the best test performance in one or several domain(s) of interest. This multi-source (multi-domain) adaptation problem is often approached with instance selection or reweighting strategies, most of which pre-supposes an ad-hoc assessment of relevance for training sentences. In this contribution, we study the recently proposed Differential Data Selection (DDS) model  and explore its ability to serve an effective generic framework for these various situations. Our experiments for both domain adaptation and multi-domain learning show that DDS often enables to outperform heuristic training strategies; we also introduce a variant that boosts DDS performance of adapter-based multi-domain systems.

\section{Learning with multiple data sources} \label{sec:mdmt}

@@

Given $\vtheta_t$, $\psi_t$ is updated via the Reinforce algorithm \cite{Williams92simple}:\fyTodo{Rewrite this correctly with domain average reward}
\begin{align*}
\psi_{t+1} &= \psi_t + \mathbf{lr}_{data} * \displaystyle{\mathop{\sum}_{d=1}^K} R(d) * \frac{\partial log(\lambda^l(d;\psi_t))}{\partial \psi} \\
  R(d) &= \langle \frac{\partial J^l_d(\theta_t)}{\partial \theta}, \frac{\partial J^t(\theta', \lambda^t)}{\partial \theta} \rangle \\
\end{align*}
Where
\begingroup
\allowdisplaybreaks
\begin{align*}
\theta' &= \theta_t - \mathbf{lr}_{nmt} * \displaystyle{\mathop{\sum}_{i=1,x^i_d,y^i_d \in \widehat{D^l_d}}^N} \frac{\partial l(\theta_t,x^i_d,y^i_d)}{\partial \theta} \\
  J^l_d(\theta_t) &= \displaystyle{\mathop{\sum}_{i=1,x^i_d,y^i_d \in \widehat{D^l_d}}^N} l(\theta_t,x^i_d,y^i_d) \\
  J^t(\theta',\lambda_t) &= \displaystyle{\mathop{\sum}_{d}}\lambda^t(d)\displaystyle{\mathop{\sum}_{x^t_d,y^t_d \in \widehat{D^t_d}}} l(\theta',x^t_d,y^t_d)\\
\end{align*}
\endgroup
$\mathbf{lr}_{data}$ is the learning rate of the parameter of the sampler. $\mathbf{lr}_{nmt}$ is the learning rate of the parameter of the NMT system. \fyTodo{Explain lr, may be more}\fyTodo{More refs to the algorithm }
% To apply this method to multi-domain training (\system{Multi-DDS}), we limit the search family of training distribution $D_{trn}$ to the family of mixture of K in-domain training distribution such as $\sum_d \lambda^{s}_{d} \mathcal{D}_d(x,y)$. We also parameterize $\lambda^{s}_{d}$ using Softmax function so that $\lambda^{s}_{d} = \text{Softmax}(\psi)_d$ where $\psi \in \mathbb{R}^K$.
\subsection{DDS for (multi) domain adaptation}

@@

\fyTodo{Fix the table}
In Table ~\ref{tab:basic-dds}, we compare the results of fine-tuning a system initially trained with a natural mixture, which is proportional to the sizes of the domains, of domains with the result of DDS learning from scratch with a initial uniform distribution. \fyTodo{What do we see ? How about domain proximity ? - show a graph}.

\fyTodo{Only if we compare on an individual domain} \revision{Note that the DDS systems are somewhat quicker to build, as we only train for 200K iterations (instead of 200K + 50K for the fine-tuning part)}.

\begin{table*}
  \centering \small
  \begin{tabular}{|l|*8{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} \\ \hline
    FT-Full($d$) \hfill [BLEU] &37.7&59.2&54.5&34&46.8&90.8&53.8\\
    FT-Res($d$) \hfill &37.3&57.9&53.9&33.8&46.7&90.2&53.3\\ \hline
    \hline
    \system{DDS($\vlambda_0, \vlambda_d$)} \hfill [BLEU] &36.7&56.2&54.3&33.5&46.8&91.8&53.2\\
    \% domain $d$ (final / total) & 59.7 / 49.5 & 6.5 / 18.3 & 14.9 / 24.5 & 6.4 / 12.8 & 67.2/47.4 & 67.2 / 45.3&\\ \hline
    \system{DDS($\vlambda_1, \vlambda_d$)} \hfill [BLEU]&36.4&56.3& 45.4& 32.6&44.5&43.1&43.1\\
    \% domain $d$& 89 / 86.6 & 50/18.4 & 1.4 / 1.6 & 3.0 / 2.5 & 7.5 / 9.2 & 1.0 / 1.0 \\ \hline
    \system{CL($d$)} \hfill [BLEU]&  \\ 
    \% domain $d$& \\ \hline\hline
    % Multidomain
    \system{Mixed-0}      & 35.3 & 54.1 & 52.5 & 31.9 & 44.9 & 89.5& 51.4 \\
    \system{Mixed-0.25} & 35.9 & 54.9 & 52.6 & 32.6 & 45 & 90.3& 51.9 \\
    \system{Mixed-0.5}   & 36.1 & 55.4 & 51.8 & 33.5 & 46.2 & 90 & 52.2 \\
    \system{Mixed-0.75} & 36.5 & 55 & 51.2 & 34 & 44.3 & 87.2& 51.7 \\
    \system{Mixed-1}     & 37.3  & 54.6 & 50.1 & 33.5 & 43.2 & 77.5& 49.4 \\
    \system{MDL-Res} &37.1&57.4&53.6&33.2&46&90.7&53\\\hline
    \system{DDS($\vlambda_1, \vlambda_0$)} &37.9&50.2&47.2&31.2&40.8&58.6&44.3\\
    $\operatorname{KL}(\vlambda_*^{l}, \vlambda_0$) &37&55.6&52.9&33.1&44.5&91.3&52.4 \\
%    $ \mathcal{H}(\vlambda_*^{l})$ &  \\
    \system{DDS($\vlambda_0, \vlambda_0$)} &37.3&55.1&51&33.5&43.4&90.8&51.8\\ \hline
  \end{tabular}
  \caption{Baseline results for DDS in fine-tuning and multi-domain adaptation. In the top part, we report scores with DDS and each target domain, and compare with full fine-tuning: each column corresponds to a distinct system. We also report the proportion of in-domain data in the final mixture. In the bottom part, we run multi-domain experiments: for a given line, all the columns \emph{correspond to the same system}. We also report the KL divergence between the optimal and the target mixtures.}
  \label{tab:basic-dds}
\end{table*}
\fyTodo{Report CL results / explain in legend}

@@

\subsection{Unsupervised DDS}
To better analyze the behavior of DDS, we study a scenario where $\vlambda^s$ and $\vlambda^t$ have different dimensions designed as follows. All training data is automatically clustered into $K=30$ of varying sizes using the k-means algorithm based on generic sentence representations obtained via mean pooling. The resulting classes contain from a tenths of thousands to several hundreds of thousand training samples. Most classes contain a dominant domain which accounts for more of 90\% of the samples.\footnote{See detail in supplementary material} Using the same DA setting as before, we study (a) whether DDS is able to identify the classes that correspond to the domain of interest (b) whether having additional degrees of freedom\fyTodo{explain better} can yield an improved training performance (c) how DDS scales when a large number of  domains is considered.\fyTodo{TBC - with what experiments?}

@@

This idea is reminiscent of the work of \citet{Graves17automated} and \citet{Kumar19reinforcement}, who also design their curriculum trainng strategy based on the development loss.

\subsection{Improving dynamic data selection}
Experiments with these revised versions are for various settings in Table~\ref{tab:improved-dds}. Results for DA directly compare with Table~{tab:basic-dds}; results for unsupervised DA used our 6 domains for training, but test on the \domain{News}, unseen in traning.
\begin{table*}
  \centering \small
  \begin{tabular}{|l|*9{r|}} \hline
    \multicolumn{9}{|c|}{\sl Supervised domain adaptation} \\ \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} & \multicolumn{1}{|c|}{\domain{news}} \\ \hline \hline
    \multicolumn{9}{|c|}{\sl Unsupervised domain adaptation} \\ \hline
    \system{DDS-v1($\vlambda_0, \vlambda_{news}$)} & \\  \hline\hline 
    \multicolumn{9}{|c|}{\sl 30 clusters} \\ \hline
    \\  \hline\hline 
    \multicolumn{9}{|c|}{\sl Multi-domain training} \\ \hline
    \\
    \system{DDS-v1($\vlambda_0, \vlambda_d$)} &36.7&57.6&54.9&33.3&46.1&91.9&53.4&- \\
    \system{DDS-v1($\vlambda_1, \vlambda_d$)} & \\
    \system{DDS-v1($\vlambda_0, \vlambda_0$)} &36.96&55.55&52.88&33.06&44.52&91.25&52.37&- \\
    \system{DDS-v1($\vlambda_1, \vlambda_0$)} & \\ \hline
  \end{tabular}
  \caption{Improved results for DDS in fine-tuning and various multi-domain adaptation settings.}
  \label{tab:basic-dds}
\end{table*}

>>>>

\title{Automated curriculum learning for Multi-Domain NMT by using loss-driven reward}

@@

\begin{document}
\maketitle
\setlength{\abovedisplayskip}{2pt}
\setlength{\belowdisplayskip}{2pt}
\begin{abstract}
  When building machine translation systems, one often needs to make the best out of heterogeneous sets of parallel data in training to achieve the best test performance in one or several domain(s) of interest. This multi-source / multi-domain adaptation problem is often approached with instance selection or reweighting strategies, most of which pre-suppose an ex-ante assessment of the relevance of training instances. In this paper, we propose a novel dynamic data selection algorithm which re-evaluates the data utility and adapts the data selection policy along the training of the NMT model. We explore the ability of our method to serve as a generic framework for the previously mentioned situations.
\end{abstract}

@@

This problem is often approached by \emph{static} instance selection or re-weighting strategies, where the available training data is used in proportion to its relevance for the chosen test conditions. Finding the optimal balance of training data is however a challenging task due, for instance, to similarity between domains / languages, but also due to regularization effects of out-of-domain data; it may also be suboptimal, as some target domains or languages might be easier to train than others. Several recent proposals \cite{Wang17instance,Zhang19curriculum,Kumar19reinforcement,Wang20learning-multi,alex17automated} have explored ways to instead consider \emph{dynamic} data selection and sampling strategies. In this paper, we propose a nouvel method of building an automated curriculum for training Multi-Domain NMT model, that is inspired by the work of \cite{alex17automated}. We evaluate our method in various single-domain and multi-domain adaptation settings. We compare our method with several contrasts including the work of \cite{Zhang19curriculum} and \cite{Wang20balancing}. From now on, we name our method MD-ACL

Based on experimental results obtained on a diverse set of domains, our main conclusions are that (a) using MD-ACL often yields overall performance that are as good as the standard fine-tuning strategy for domain adaptation; (b) MD-ACL can effectively handle a variety of test target distributions without any meta-parameter search.
%(c) our extension of DDS is able to boost its performance when used in conjunction to the adapter model of \citet{Bapna19simple}.
\section{Learning with multiple data sources} \label{sec:mdmt}

@@

Given $\vtheta_t$, $\psi_t$ is updated as follows.
%via the Reinforce algorithm \cite{Williams92simple}:\fyTodo{Rewrite this correctly with domain average reward}
\begin{align*}
\psi_{t+1} &= \psi_t + \mathbf{lr}_{data} * \displaystyle{\mathop{\sum}_{d=1}^K} R(d) * \frac{\partial log(\lambda^l(d;\psi_t))}{\partial \psi} \\
  R(d) &= \langle \frac{\partial J^l_d(\theta_t)}{\partial \theta}, \frac{\partial J^t(\theta', \lambda^t)}{\partial \theta} \rangle \\
\end{align*}
Where
\begingroup
\allowdisplaybreaks
\begin{align*}
\theta' &= \theta_t - \mathbf{lr}_{nmt} * \displaystyle{\mathop{\sum}_{i=1,x^i_d,y^i_d \in \widehat{D^l_d}}^N} \frac{\partial l(\theta_t,x^i_d,y^i_d)}{\partial \theta} \\
  J^l_d(\theta_t) &= \displaystyle{\mathop{\sum}_{i=1,x^i_d,y^i_d \in \widehat{D^l_d}}^N} l(\theta_t,x^i_d,y^i_d) \\
  J^t(\theta',\lambda_t) &= \displaystyle{\mathop{\sum}_{d}}\lambda^t(d)\displaystyle{\mathop{\sum}_{x^t_d,y^t_d \in \widehat{D^t_d}}} l(\theta',x^t_d,y^t_d)\\
\end{align*}
\endgroup
$\mathbf{lr}_{data}$ is the learning rate of the parameter of the sampler. $\mathbf{lr}_{nmt}$ is the learning rate of the parameter of the NMT system. \fyTodo{Explain lr, may be more}\fyTodo{More refs to the algorithm }
% To apply this method to multi-domain training (\system{Multi-DDS}), we limit the search family of training distribution $D_{trn}$ to the family of mixture of K in-domain training distribution such as $\sum_d \lambda^{s}_{d} \mathcal{D}_d(x,y)$. We also parameterize $\lambda^{s}_{d}$ using Softmax function so that $\lambda^{s}_{d} = \text{Softmax}(\psi)_d$ where $\psi \in \mathbb{R}^K$.
\subsection{DDS for (multi) domain adaptation}

@@

\fyTodo{Fix the table}
In Table ~\ref{tab:basic-dds}, we compare the results of fine-tuning a system initially trained with a natural mixture, which is proportional to the sizes of the domains, of domains with the result of DDS learning from scratch with a initial uniform distribution.

\begin{table*}
  \centering \small
  \begin{tabular}{|l|*8{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} \\ \hline
    FT-Full($d$) \hfill [BLEU] &37.7&59.2&54.5&34&46.8&90.8&53.8\\
    FT-Res($d$) \hfill &37.3&57.9&53.9&33.8&46.7&90.2&53.3\\ \hline
    \hline
    \system{DDS($\vlambda_0, \vlambda_d$)} \hfill [BLEU] &36.7&56.2&54.3&33.5&46.8&91.8&53.2\\
    \% domain $d$ (final / total) & 59.7 / 49.5 & 6.5 / 18.3 & 14.9 / 24.5 & 6.4 / 12.8 & 67.2/47.4 & 67.2 / 45.3&\\ \hline
    \system{DDS($\vlambda_1, \vlambda_d$)} \hfill [BLEU]&36.4&56.3& 45.4& 32.6&44.5&43.1&43.1\\
    \% domain $d$& 89 / 86.6 & 50/18.4 & 1.4 / 1.6 & 3.0 / 2.5 & 7.5 / 9.2 & 1.0 / 1.0 \\ \hline
    \system{CL($d$)} \hfill [BLEU]&  \\ 
    \% domain $d$& \\ \hline\hline
    % Multidomain
    \system{Mixed-0}      & 35.3 & 54.1 & 52.5 & 31.9 & 44.9 & 89.5& 51.4 \\
    \system{Mixed-0.25} & 35.9 & 54.9 & 52.6 & 32.6 & 45 & 90.3& 51.9 \\
    \system{Mixed-0.5}   & 36.1 & 55.4 & 51.8 & 33.5 & 46.2 & 90 & 52.2 \\
    \system{Mixed-0.75} & 36.5 & 55 & 51.2 & 34 & 44.3 & 87.2& 51.7 \\
    \system{Mixed-1}     & 37.3  & 54.6 & 50.1 & 33.5 & 43.2 & 77.5& 49.4 \\
    \system{MDL-Res} &37.1&57.4&53.6&33.2&46&90.7&53\\\hline
    \system{DDS($\vlambda_1, \vlambda_0$)} &37.9&50.2&47.2&31.2&40.8&58.6&44.3\\
    $\operatorname{KL}(\vlambda_*^{l}, \vlambda_0$) &37&55.6&52.9&33.1&44.5&91.3&52.4 \\
%    $ \mathcal{H}(\vlambda_*^{l})$ &  \\
    \system{DDS($\vlambda_0, \vlambda_0$)} &37.3&55.1&51&33.5&43.4&90.8&51.8\\ \hline
  \end{tabular}
  \caption{Baseline results for DDS in fine-tuning and multi-domain adaptation. In the top part, we report scores with DDS and each target domain, and compare with full-fine-tuning: each column corresponds to a different system. We also report the proportion of in-domain data in the final mixture. In the bottom part, we run multi-domain experiments: for a given line, all the columns \emph{correspond to the same system}. We also report the KL divergence between the optimal and the target mixtures.}
  \label{tab:basic-dds}
\end{table*}

@@

\subsection{Unsupervised DDS}
To better analyze the behavior of DDS, we study a scenario where $\vlambda^s$ and $\vlambda^t$ have different dimensions designed as follows. All training data is automatically clustered into $K=30$ of varying sizes using the k-means algorithm based on generic sentence representations obtained via mean pooling. The resulting classes contain from a tenths of thousands to several hundreds of thousand training samples. Most classes contain a dominant domain, which accounts for more of 90\% of the samples.\footnote{See detail in supplementary material} Using the same DA setting as before, we study (a) whether DDS is able to identify the classes that correspond to the domain of interest (b) whether having additional degrees of freedom can yield an improved training performance (c) how DDS scales when a large number of  domains is considered.\fyTodo{TBC - with what experiments?}

@@

This idea is reminiscent of the work of \citet{alex17automated} and  \citet{Kumar19reinforcement}, who also design their curriculum trainng strategy based on the development loss.

\subsection{Improving dynamic data selection}
Experiments with these revised versions are for various settings in Table~\ref{tab:improved-dds}. Results for DA directly compare with Table~{tab:basic-dds}; results for unsupervised DA used our 6 domains for training, but test on the \domain{News}, unseen in traning.
\begin{table*}
  \centering \small
  \begin{tabular}{|l|*9{r|}} \hline
    \multicolumn{9}{|c|}{\sl Supervised domain adaptation} \\ \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} & \multicolumn{1}{|c|}{\domain{news}} \\ \hline \hline
    \multicolumn{9}{|c|}{\sl Unsupervised domain adaptation} \\ \hline
    \system{DDS-v1($\vlambda_0, \vlambda_{news}$)} & \\  \hline\hline 
    \multicolumn{9}{|c|}{\sl 30 clusters} \\ \hline
    \system{DDS-v1($\vlambda^*, \vlambda_d$)}& &56.9&51.3&33.2&43.4&&&
    \\  \hline\hline 
    \multicolumn{9}{|c|}{\sl Multi-domain training} \\ \hline
    \\
    \system{DDS-v1($\vlambda_0, \vlambda_d$)} &36.7&58.2&55&33.3&46.6&91.9&53.6& 23.1 \\
    \system{DDS-v1($\vlambda_1, \vlambda_d$)} & \\
    \system{DDS-v1($\vlambda_0, \vlambda_0$)}&35.5&55.5&53.7&32.3&45.6&91.1&52.3& - \\
    \system{DDS-v1($\vlambda_1, \vlambda_0$)}&37.4&54.4&50.3&32.0&42.7&80.2&49.5 \\ \hline
  \end{tabular}
  \caption{Improved results for DDS in fine-tuning and various multi-domain adaptation settings.}
  \label{tab:basic-dds}
\end{table*}
